---
title: "Using `vcd2df` to load a Value Change Dump as a Data Frame"
author: "Calvin Deutschbein"
description: >
 This document provides an example usage of the `vcd2df` function.
vignette: >
  %\VignetteIndexEntry{index}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format: 
  html:
    theme: none
    minimal: true
    embed-resources: true
---

## Value Change Dump files

We provide an example of the `vcd2df` function loading a IEEE 1364-2001 
VCD (.vcd) file, specified as an parameter as a string containing exactly 
a file path, and returns an R dataframe containing values over time.
  
A VCD file captures the register values at discrete timepoints from a 
simulated trace of execution of a hardware design in verilog or VHDL.

The returned dataframe contains a row for each register, by name, and a
column for each time point, specified VCD-style using octothorpe-prefixed 
multiples of the timescale. 

The only non-trivial implementation details are that

1. VCD 'x' and 'z' non-numerical values are encoded as negativei values 
(as otherwise all bit values are positive) and 
2. registers with repeated names in distinct modules are ignored, rather 
than duplicated, as we anticipate these registers to have the same values.

## Setup 

The library and function share the same name, which we load as follows:


```{r}
#| label: setup
library(vcd2df)
```

## Sample Data

We have created a [small repository](https://github.com/vcd2df/vcd_ex) with a
few VCD files from common development platforms. We will download a 
VCD file from the smallest, the "Naive Educational RISC-V Processor"
(nerv) which uses a short testbench and generates a manageable file.

```{r}
#| label: download
f_name <- tempfile()
vcd <- 'https://github.com/vcd2df/vcd_ex/raw/refs/heads/main/nerv.vcd'
download.file(url = vcd, destfile = f_name)
```

The file `nerv.vcd` is created as specified in the `Makefile` of the
[YosysHQ/nerv](https://github.com/YosysHQ/nerv) repository by running
`make test` which generates a VCD file called `testbench.vcd`. We simply
renamed this file to differentiate to document it's source design.

## Load into data frame

With a local VCD file, loading is a single function call.

```{r}
#| label: read
df <- vcd2df(f_name)
```

### File Management

Once loaded within a data frame, the intial file may be cleaned.

```{r}
#| label: clean
unlink(f_name)
```

## Inspect data frame

Most commonly, we can look at some register names with `rownames()`... 
```{r}
#| label: rownames
rownames(df)[1:5]
```


... or timestamps with `colnames()`.

```{r}
#| label: colnames
colnames(df)[1:5]
```

From there, the entire R data science workflow is accessible.
